Efficient evaluation of threshold queries of derived fields in a numerical simulation database

### 2014-09-03 paper.tex
Work on Conclusions section.

@@

\section{Scientific Use Cases}\label{science_use_cases}
%Max vorticity clustering.
One of the applications of thresholding in turbulence is to find the locations of maximum vorticity in a particular time-step or the dataset as a whole.
The locations of maximum vorticity are usually associated with the most intense vortices in the dataset and often have interesting and complex
reconnection events associated with them. 
Once obtained from the service, these locations can be clustered in both 3d and 4d. This allows scientists to examine their evolution with the flow
and make subsequent analysis queries as needed in order to study these events.

@@

\section{JHTDB}
The method that we have developed for the evaluation of threshold queries of derived fields was deployed and integrated into the 
Johns Hopkins Turbulence Databases (JHTDB). The JHTDB, built on top of the GrayWulf and DataScope clusters, serve as a public virtual laboratory for
the study of turbulent phenomena. The JHTDB store several datasets, which are the output of high-resolution numerical simulations of turbulence.
The 3d time-series data are partitioned into small sub-cubes and stored in relational databases distributed across the nodes of the cluster.
Access to the data is provided by means of Web-services and a variety of analysis functions have been
implemented and can be executed through Web-service calls.
At present the service hosts four datasets, which are available publicly. 
The data are the output of numerical simulations of forced isotropic turbulence, magnetohydrodynamics (MHD), channel 
flow turbulence and homogenous buoyancy driven turbulence.
The total amount of space occupied by the datasets is over 230 TB.

@@

For this evaluation we looked at the performance of threshold queries to the vorticity field. The vorticity field is representative of derived fields, which
have to be computed from the stored data. The vorticity is computed from the velocity field by taking its curl:
%\begin{align}
%&\sum_{i_0=0}^{1} \sum_{i_1=0}^{1} ... \sum_{i_{d-1}=0}^{1} {(-1)}^{i_0 + i_1 + ...+ i_{d-1}}\cdot  \nonumber \\
%&\qquad s(x_0(i_0), x_1(i_1), ..., x_{d-1}(i_{d-1}))
%\label{eq:summed_volume}
%\end{align}
\begin{align}
\vec{\omega} &= \vec{\nabla} \times \vec{v} = \left( \frac{\partial}{\partial x}, \frac{\partial}{\partial y}, \frac{\partial}{\partial z} \right) 
\times \nonumber \left( v_x, v_y, v_z \right)\\
&= \left( \frac{\partial v_z}{\partial y} - \frac{\partial v_y}{\partial z}, \frac{\partial v_x}{\partial z} - \frac{\partial v_z}{\partial x}, 
\frac{\partial v_y}{\partial x} - \frac{\partial v_x}{\partial y} \right).
\label{eq:curl}
\end{align}
We use finite differencing methods of
different orders for the evaluation of the curl. For example, with 4th-order centered finite differencing each partial derivative is evaluated from the 4
adjacent grid node values as follows:  
%\begin{align}
%\frac{df}{dx}
%\label{eq:curl}
%\end{align}
\begin{align} 
\left.\frac{df}{dx}\right|_{x_n} = &\frac{2}{3\Delta x}[f(x_{n+1})-f(x_{n-1})]\nonumber\\ 
&-\frac{1}{12\Delta x}[f(x_{n+2})-f(x_{n-2})],
\end{align}
where $f$ denotes any one of the three components of the velocity and $\Delta x$ is the width of the grid in the $x$ direction.
To illustrate the distribution of the values for the norm of the vorticity field in the MHD dataset and to get an appreciation for the different threshold values
used in the experiments we show a coarse view of the probability density function for the vorticity in Fig. \ref{fig:pdf}.  
As described in section \ref{science_use_cases}
thresholding the vorticity field is important in the study of fluid dynamics and clustering the locations of maximum vorticity can lead to new insights into
the development of the most intense vortices that are observed. 
%\kk{Consider showing other fields as well; possibly thresholding the velocity by itself}

\subsection{Evaluation of cache effectiveness}
\begin{table}[h]
\centering
\begin{tabular}{|c|c|c|c|c|} \hline
\multirow{2}{0.5in}{Threshold} & \multirow{2}{0.5in}{Points above threshold} & \multicolumn{3}{|c|}{Running time (s.)}\\ \cline{3-5}
& & No cache & \parbox[t]{0.55in}{With cache (miss)} & \parbox[t]{0.55in}{With cache (hit)} \\ \hline
80.0 & 4247 & 97.1 & 100.2 & 0.5\\ \hline
60.0 & 86580 & 113.7 & 115.9 & 1.2\\ \hline
44.0 & 909274 & 111.6 & 115.0 & 9.1\\ \hline
\end{tabular}
\caption{Effectiveness of caching.}
\label{tab:cache_effectiveness}
\end{table}
The central part of the strategy that we have developed for the evaluation of threshold queries of derived fields is the application-aware cache, which
stores the results of these queries. We first evaluate the overhead associated with the introduction and maintenance of the cache. Fig. \ref{fig:cache_overhead}
compares the execution time of queries in the absence of a cache with the execution time of the same queries, which interrogate the cache first.
For these experiments we requested
the locations with norms of the vorticity above thresholds at different levels. For the first set the threshold was set high (80.0) and only $\sim$4,300 points 
(or 0.0004\% of all points) were above the threshold. For the second set a medium threshold (60.0) was chosen and $\sim$87,000 points 
(or 0.0081\% of all points) were above the threshold. Finally, a low threshold (44.0) was chosen for the last set with $\sim$900,000 points 
(or 0.0847\% of all points) above the threshold. For each set a random
time-step was chosen and the queries were run against that time-step. The measurements were taken from the point of view of the end user. 
The cache was initially populated by executing several hundred unrelated queries and contained several million entries.
Cache entries for the particular time-step queried were dropped before each run, making sure that each query would produce a cache miss and would
have to be evaluated from the raw data. 
The execution times were averaged over 10 runs.
We utilized 4 threads per database
node for the evaluation of each query. As we can see from the results the overhead associated with querying the cache first is minimal, 
less than 3\% and within
the margin of error. Additionally, the method shows stable running time across different time-steps and threshold levels. The slight increase in the running
time that can be observed is due to the larger result set that has to be returned to the user.

@@

\subsection{Scaling and Distributed Evaluation}
The data-parallel evaluation of threshold queries exhibits good vertical scaling as shown in Fig. \ref{fig:scaleup}. For these experiments we used the same
queries and threshold values as for the runs shown in Fig. \ref{fig:cache_overhead} but with varying number of threads per node. Cache entries for the
time-step queried were again dropped before each run in order to evaluate the scaling properties of the computation of the derived field from the stored data.
The computations for all of 
the derived fields of interest (such as the vorticity) at each gird point need data from adjacent grid points only. Therefore, each node of the cluster is able to
compute the derived field from data available locally with only a little amount along the boundary of each region having to be retrieved from adjacent nodes.
Each computation is independent and can be executed in parallel. This allows us to make use of multiple threads per node.

@@

- Lays the ground work for the creation of a landmark database, which could store the locations of the highest vorticity regions in the dataset.

@@

>>>>

\section{Scientific Use Cases}\label{science_use_cases}

%Max vorticity clustering.
One of the applications of thresholding in turbulence is to find the locations of maximum vorticity in a particular time-step or the dataset as a whole.
The locations of maximum vorticity are usually associated with the most intense vortices in the dataset and often have interesting and complex
reconnection events associated with them. 
Once obtained from the service, these locations can be clustered in both 3d and 4d. This allows scientists to examine their evolution with the flow
and make subsequent analysis queries as needed in order to study these events.

The vorticity is computed from the velocity field by taking its curl:
\begin{align}
\vec{\omega} &= \vec{\nabla} \times \vec{v} = \left( \frac{\partial}{\partial x}, \frac{\partial}{\partial y}, \frac{\partial}{\partial z} \right) 
\times \nonumber \left( v_x, v_y, v_z \right)\\
&= \left( \frac{\partial v_z}{\partial y} - \frac{\partial v_y}{\partial z}, \frac{\partial v_x}{\partial z} - \frac{\partial v_z}{\partial x}, 
\frac{\partial v_y}{\partial x} - \frac{\partial v_x}{\partial y} \right).
\label{eq:curl}
\end{align}
We use finite differencing methods of
different orders for the evaluation of the curl. For example, with 4th-order centered finite differencing each partial derivative is evaluated from the 4
adjacent grid node values as follows:  
\begin{align} 
\left.\frac{df}{dx}\right|_{x_n} = &\frac{2}{3\Delta x}[f(x_{n+1})-f(x_{n-1})]\nonumber\\ 
&-\frac{1}{12\Delta x}[f(x_{n+2})-f(x_{n-2})],
\end{align}
where $f$ denotes any one of the three components of the velocity and $\Delta x$ is the width of the grid in the $x$ direction. The partial derivatives
along $y$ and $z$ are computed in the same fashion. 
Fig. \ref{fig:pdf} shows the distribution of the values of the norm of the vorticity field in the MHD dataset for a representative time-step. This is indicative 
of how the values are distributed in the dataset as a whole.
This coarse view of the data can be used by scientists to guide the selection of threshold values.

@@

@@

\section{JHTDB}

The method that we have developed for the evaluation of threshold queries of derived fields was deployed and integrated into the 
Johns Hopkins Turbulence Databases (JHTDB). The JHTDB, built on top of the GrayWulf and DataScope clusters, serve as a public virtual laboratory for
the study of turbulent phenomena. The JHTDB store several datasets, which are the output of high-resolution numerical simulations of turbulence.
The 3d time-series data are partitioned into small sub-cubes and stored in relational databases distributed across the nodes of the cluster.
Access to the data is provided by means of Web-services and a variety of analysis functions have been
implemented and can be executed through Web-service calls.
At present the service hosts four datasets, which are available publicly. 
The data are the output of numerical simulations of forced isotropic turbulence, magnetohydrodynamics (MHD), channel 
flow turbulence and homogenous buoyancy driven turbulence.
The total amount of space occupied by the datasets is over 230 TB.

@@

For this evaluation we looked at the performance of threshold queries to the vorticity field. The vorticity field is representative of derived fields, which
have to be computed from the stored data. It defined as the curl of the velocity field.
As described in section \ref{science_use_cases}
thresholding the vorticity field is important in the study of fluid dynamics and obtaining the locations of maximum vorticity can lead to new insights into
the development of the most intense vortices that are observed. 
%\kk{Consider showing other fields as well; possibly thresholding the velocity by itself}

\subsection{Evaluation of cache effectiveness}

The central part of the strategy that we have developed for the evaluation of threshold queries of derived fields is the application-aware cache, which
stores the results of these queries. We first evaluate the overhead associated with the introduction and maintenance of the cache. Fig. \ref{fig:cache_overhead}
compares the execution time of queries in the absence of a cache with the execution time of the same queries, which interrogate the cache first.
For these experiments we requested
the locations with norms of the vorticity above thresholds at different levels. We refer the reader to Fig. \ref{fig:pdf}, which shows the distribution of values of
the norm of the vorticity field in the MHD dataset to get an appreciation of the different threshold values used in the experiments. 
For the first set the threshold was set high (80.0) and only $\sim$4,300 points 
(or 0.0004\% of all points) were above the threshold. For the second set a medium threshold (60.0) was chosen and $\sim$87,000 points 
(or 0.0081\% of all points) were above the threshold. Finally, a low threshold (44.0) was chosen for the last set with $\sim$900,000 points 
(or 0.0847\% of all points) above the threshold. For each set a random
time-step was chosen and the queries were run against that time-step. The measurements were taken from the point of view of the end user.

The cache was initially populated by executing several hundred unrelated queries and contained several million entries.
Cache entries for the particular time-step queried were dropped before each run, making sure that each query would produce a cache miss and would
have to be evaluated from the raw data. 
The execution times were averaged over 10 runs.
We utilized 4 threads per database
node for the evaluation of each query. As we can see from the results the overhead associated with querying the cache first is minimal, 
less than 3\% and within
the margin of error. Additionally, the method shows stable running time across different time-steps and threshold levels. The slight increase in the running
time that can be observed is due to the larger result set that has to be returned to the user.

\begin{table}[h]
\centering
\begin{tabular}{|c|c|c|c|c|} \hline
\multirow{2}{0.5in}{Threshold} & \multirow{2}{0.5in}{Points above threshold} & \multicolumn{3}{|c|}{Running time (s.)}\\ \cline{3-5}
& & No cache & \parbox[t]{0.55in}{With cache (miss)} & \parbox[t]{0.55in}{With cache (hit)} \\ \hline
80.0 & 4247 & 97.1 & 100.2 & 0.5\\ \hline
60.0 & 86580 & 113.7 & 115.9 & 1.2\\ \hline
44.0 & 909274 & 111.6 & 115.0 & 9.1\\ \hline
\end{tabular}
\caption{Effectiveness of caching.}
\label{tab:cache_effectiveness}
\end{table}

@@

\subsection{Scaling and Distributed Evaluation}
The data-parallel evaluation of threshold queries exhibits good vertical scaling as shown in Fig. \ref{fig:scaleup}. For these experiments we used the same
queries and threshold values as for the runs shown in Fig. \ref{fig:cache_overhead} and Table \ref{tab:cache_effectiveness} 
but with varying number of threads per node. Cache entries for the
time-step queried were again dropped before each run in order to evaluate the scaling properties of the computation of the derived field from the stored data.
The computations for all of 
the derived fields of interest (such as the vorticity) at each gird point need data from adjacent grid points only. Therefore, each node of the cluster is able to
compute the derived field from data available locally with only a little amount along the boundary of each region having to be retrieved from adjacent nodes.
Each computation is independent and can be executed in parallel. This allows us to make use of multiple threads per node.

@@

We have presented an efficient strategy for the evaluation of threshold queries of derived fields in large numerical simulation datasets. 
The fields, which are thresholded are derived from the stored simulation data in a distributed data-parallel manner. The computations
scale with the cluster resources and are performed on the database nodes, where the data are stored. This new capability allows researches to quickly
obtain and focus on regions of special interest even if they lack the computing capabilities or data transfer rates necessary to examine entire time-steps
or large parts of the entire dataset.

We have introduced an application-aware cache for the query results. The cache adds minimal overhead during the evaluation of queries even if there is
a cache miss and has modest storage requirements. Cache hits reduce query running times by over an order of magnitude.

The introduction of an application-aware cache for query results lays the ground work for the creation of a landmark database. Such a database can store the 
locations of the highest vorticity regions in the dataset or more broadly regions of interest and their associated statistics.

The Web-services approach to archived numerical simulation datasets provides public access to high quality simulation data to anyone with an internet
connection. The Web-services methods can be called from any modern programming language and we provide C, Fortran and Matlab client libraries for the
JHTDB. The evaluation of each query submitted through a Web-service call is carried out on the nodes of the database cluster by mens of a stored procedure
or a user-defined function that has been implemented and deployed to handle these requests. This allows us to fine-tune the execution of these procedures
and handle all requests transparently to the user. However, this approach also has drawbacks. Adding new functionality means adding to a long list of 
Web-service calls and requires substantial implementation effort. In the case of threshold queries the stored procedure performing the evaluation must have
an implementation for each derived field of interest even though the execution is handled by the same Web-service call.

In the future we plan to develop
declarative and graphical user interfaces that will allow users to combine existing building blocks and perform computations that have not been explicitly
implemented. Additionally, we plan on deploying a server-side computing environment for users similar to the CasJobs service for the Sloan Digital Sky Survery \cite{LiThakar}. In such an environment users can run queries in batch mode and save their results in a personal database called MyDB, which 
will allow for much greater flexibility in the type of computations that can be performed in addition to substantially decreasing the network overhead.

@@

The authors would like to thank the Turbulence Database Group at Johns Hopkins University for
their insightful comments and suggestions. This work is supported in part by the National Science Foundation
under Grants ....
